---
title: "Exploratory Data Analysis"
author: "ak5357"
date: "2024-10-03"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(dplyr)
library(haven)
```


## Import weather data.
```{r import_weather_data}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2021-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = case_match(
      id, 
      "USW00094728" ~ "CentralPark_NY", 
      "USW00022534" ~ "Molokai_HI",
      "USS0023B17S" ~ "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10,
    month = lubridate::floor_date(date, unit = "month")) |>
  select(name, id, everything())
```

## Exploratory visuals
```{r weather_hist}
weather_df |> 
  ggplot(aes(x = prcp)) + 
  geom_histogram()
```

**Filter by precipitation.**
```{r weather_low_prcp}
weather_df |> 
  filter(prcp >= 1000)
```

**Plot weather scatterplot.**
```{r weather_plot_1}
weather_df |> 
  filter(tmax >= 20, tmax <= 30) |> 
  ggplot(aes(x = tmin, y = tmax, color = name, shape = name)) + 
  geom_point(alpha = .75)
```


## `group_by()`
```{r weather_group_by}
# Group by name and get n obs per name
weather_df |> 
  group_by(name) |> 
  summarize(n_obs = n())

# Group by month and get n obs per name
weather_df |> 
  group_by(month) |> 
  summarize(n_obs = n())

# Group by name and get n obs per name & n distinct months with measures
weather_df |> 
  group_by(name) |> 
  summarize(
    n_obs = n(),
    n_dist = n_distinct(month))

# Group by both name and month and get count for each
weather_df |> 
  group_by(name, month) |> 
  summarize(n_obs = n())
```


**Count without grouping.**
```{r weather_counting}
weather_df |>
  count(month, name = "n_obs")
```


## 2x2
```{r weather_2x2}
weather_df |> 
  drop_na(tmax) |> 
  mutate(
    cold = case_when(
      tmax < 5 ~ "cold",
      tmax >= 5 ~ "not cold"
    )
  ) |> 
  group_by(name, cold) |> 
  summarize(count = n())
```

**Janitor Tabyl.**
```{r janitor_tabyl}
weather_df |> 
  drop_na(tmax) |> 
  mutate(
    cold = case_when(
      tmax < 5 ~ "cold",
      tmax >= 5 ~ "not_cold"
    )
  ) |> 
  janitor::tabyl(name, cold)
```

**Other useful summaries.**
```{r weather_summary_1}
weather_df |> 
  group_by(name) |> #can group by name, month, or both
  summarize(
    mean_tmax = mean(tmax, na.rm = TRUE),
    median_tmin = median(tmin, na.rm = TRUE),
    sd_prcp = sd(prcp, na.rm = TRUE)
  )
```

**Summarize then plot.**
```{r weather_summary_plot_1}
weather_df |> 
  group_by(name, month) |> #can group by name, month, or both
  summarize(
    mean_tmax = mean(tmax, na.rm = TRUE),
    median_tmin = median(tmin, na.rm = TRUE),
    sd_prcp = sd(prcp, na.rm = TRUE)
  ) |> 
  ggplot(aes(x = month, y = mean_tmax, color = name)) +
  geom_point() +
  geom_line()
```

**Format for readers, helpful despite "untidy"-ness.**
```{r weather_summary_kable}
weather_df |> 
  group_by(name, month) |> #can group by name, month, or both
  summarize(
    mean_tmax = mean(tmax, na.rm = TRUE)
  ) |> 
  pivot_wider(
    names_from = name,
    values_from = mean_tmax
  ) |> 
  knitr::kable(
    digits = 1,
    col.names = c("Month", "Central Park", "Molokai", "Waterhole")
    )
```

**Group, mutate, and plot.**
```{r weather_summary_plot_2}
weather_df |> 
  group_by(name) |> 
  mutate(
    mean_tmax = mean(tmax, na.rm = TRUE),
    centered_tmax = tmax - mean_tmax
  ) |> 
  ggplot(aes(x = date, y = centered_tmax, color = name)) + 
    geom_point(alpha = 0.5)
```


## Window Functions

**Find hottest / coldest days.**
```{r hot_cold_days}
# 10 coldest days in the whole dataset
weather_df |> 
  mutate(
    temp_rank = min_rank(tmax) #ranks all temperatures
  ) |> 
  filter(temp_rank < 10) #filter for 10 coldest days

# 3 coldest days for each location
weather_df |> 
  group_by(name) |> 
  mutate(
    temp_rank = min_rank(desc(tmax)) #ranks all temperatures
  ) |> 
  filter(temp_rank < 4) #filter for 3 coldest days per location

# Filter by rank without creating new column
weather_df |> 
  group_by(name) |> 
  filter(min_rank(tmax) < 4) |> 
  arrange(tmax) #keeps in mind the grouping structure, I think??
```

## Lag functions
**Make sure data is ordered and no dates are missing beforehand, or the function won't work as expected.**
```{r weather_lags}
# Lagged temps and change in temp from one day to next
weather_df |>
  group_by(name) |>
  mutate(
    lagged_temp = lag(tmax), #one day behind, or the row above
    temp_change = tmax - lag(tmax)
  ) |> 
  filter(min_rank(temp_change) < 3)

# Use summarize on the lag and lag-based columns
weather_df |>
  group_by(name) |>
  mutate(
    lagged_temp = lag(tmax), #one day behind, or the row above
    temp_change = tmax - lag(tmax)
  ) |> 
  summarize(
    temp_change_sd = sd(temp_change, na.rm = TRUE),
    temp_change_max = max(temp_change, na.rm = TRUE)
  )
```

## Learning assessments

**Learning Assessment:** In the PULSE data, the primary outcome is BDI score; it’s observed over follow-up visits, and we might ask if the typical BDI score values are roughly similar at each. Try to write a code chunk that imports, cleans, and summarizes the PULSE data to examine the mean and median at each visit. Export the results of this in a reader-friendly format.
```{r learning_assessment_1}
pulse_df = read_sas("data/public_pulse_data.sas7bdat") |> 
  janitor::clean_names() |> 
  pivot_longer(
    starts_with("bdi_score"),
    names_to = "visit",
    names_prefix = "bdi_score_",
    values_to = "bdi"
  ) |> 
  relocate(id, visit) |> 
  mutate(
    visit = replace(visit, visit == "bl", "00m"),
    visit = factor(visit, levels = str_c(c("00", "01", "06", "12"), "m"))
  ) |> 
  arrange(id, visit)

pulse_df |> 
  group_by(visit) |> 
  summarize(
    mean_bdi = mean(bdi, na.rm = TRUE),
    median_bdi = median(bdi, na.rm = TRUE)
  ) |> 
  knitr::kable(digits = 3)
```

**Learning Assessment:** In the FAS data, there are several outcomes of interest; for now, focus on post-natal day on which a pup is able to pivot. Two predictors of interest are the dose level and the day of treatment. Produce a reader-friendly table that quantifies the possible associations between dose, day of treatment, and the ability to pivot.
```{r learning_assessment_2}
# Import litter data
litter_df = 
  read_csv("./data/FAS_litters.csv", na = c("NA", " ", "", ".")) |>
  janitor::clean_names() |>
  separate(group, into = c("dose", "tx_day"), sep = 3)

# Import pup data
pup_df = 
  read_csv("./data/FAS_pups.csv", na = c("NA", " ", "", ".")) |>
  janitor::clean_names() |>
  mutate(sex = recode(sex, `1` = "male", `2` = "female")) 

# Join litter and pup data
fas_df = left_join(pup_data, litter_data, by = "litter_number") 

# Summary table
fas_df |> 
  drop_na(dose) |> 
  group_by(dose, tx_day) |> 
  summarize(
    mean_pivot = mean(pd_pivot, na.rm = TRUE)
  ) |> 
  pivot_wider(
    names_from = dose,
    values_from = mean_pivot
  ) |> 
  knitr::kable()
```