---
title: "Visualization 1"
author: "ak5357"
date: "2024-09-26"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggridges)
```


## Import data

```{r}
weather_df = 
  rnoaa::meteo_pull_monitors(
    c("USW00094728", "USW00022534", "USS0023B17S"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2021-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = case_match(
      id, 
      "USW00094728" ~ "CentralPark_NY", 
      "USW00022534" ~ "Molokai_HI",
      "USS0023B17S" ~ "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10) |>
  select(name, id, everything())


weather_df
```

## Making our first plot :-)

```{r}
ggplot(weather_df, aes(x = tmin, y = tmax)) +
  geom_point()
```


```{r}
weather_df |> 
  ggplot(aes(x = tmin, y = tmax)) +
  geom_point()
```


```{r}
ggp_weather_scatterplot = 
  weather_df |> 
  ggplot(aes(x = tmin, y = tmax)) +
  geom_point()

ggp_weather_scatterplot
```


Check why some rows are missing.

```{r}
weather_df |> 
  filter(is.na(tmax))

```


## Fancier scatterplots!

```{r}
weather_df |> 
  ggplot(aes(x = tmin, y = tmax, color = name)) + #can color by some variable
  geom_point(alpha = .3, size = .8) + #alpha makes transparent
  geom_smooth(se = FALSE) #gives a smooth curve through the scatterplot points, se = standard error bars
```

All parameters in the initial `ggplot` applies to all following lines. So when `color = name` is in the `geom_point`, then `geom_smooth` doesn't treat the three colors as different sets, but rather creates one curve for the entire dataset.
```{r}
weather_df |> 
  ggplot(aes(x = tmin, y = tmax)) +
  geom_point(aes(color = name), alpha = .3, size = .8) +
  geom_smooth(se = FALSE)
```

## Using faceting

```{r}
weather_df |> 
  ggplot(aes(x = tmin, y = tmax, color = name)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(se = FALSE) +
  facet_grid(. ~ name) #the '.' means 'everything'
```


## Other tools

Let's make a somewhat more interesting scatterplot.

```{r}
weather_df |> 
  ggplot(aes(x = date, y = tmax, color = name, size = prcp)) +
  geom_point(alpha = 0.3) +
  geom_smooth(se = FALSE, color = 'black', size = 0.7) +
  facet_grid(. ~ name)

```

## Learning assessment

```{r}
weather_df |> 
  filter(name == "CentralPark_NY") |> 
  mutate(
    tmax_fahr = (tmax * 1.8) + 32,
    tmin_fahr = (tmin * 1.8) + 32
  ) |> 
  ggplot(aes(x = tmin_fahr, y = tmax_fahr)) +
  geom_point(alpha = 0.3, color = 'cornflowerblue') +
  geom_smooth(method = "lm", se = FALSE, color = 'darkblue')

```




